---
title: "Data Science and Data Analytics (WS 2025/26)"
subtitle: "International Business Management (B. A.)"
author: 
    name: "© Benjamin Gross"
    email: benjamin.gross@ext.hs-fresenius.de
    affiliation: 
      - Hochschule Fresenius - University of Applied Science
      - "Email: benjamin.gross@ext.hs-fresenius.de"
      - "Website: https://drbenjamin.github.io"
filters:
  - scripts/adjustTime.lua
abstract: |
  This document provides the course material for Data Science and Data Analytics (B. A. – International Business Management). Upon successful completion of the course, students will be able to: recognize important technological and methodological advancements in data science and distinguish between descriptive, predictive, and prescriptive analytics; demonstrate proficiency in classifying data and variables, collecting and managing data, and conducting comprehensive data evaluations; utilize R for effective data manipulation, cleaning, visualization, outlier detection, and dimensionality reduction; conduct sophisticated data exploration and mining techniques (including PCA, Factor Analysis, and Regression Analysis) to discover underlying patterns and inform decision-making; analyze and interpret causal relationships in data using regression analysis; evaluate and organize the implementation of a data analysis project in a business environment; and communicate the results and effects of a data analysis project in a structured way.
format:
  html:
    embed-resources: true
    theme: cerulean
    toc: true
    toc-expand: 5
    toc-depth: 5
    number-sections: false
    number-depth: 5
  pdf:
    keep-tex: false
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    documentclass: scrartcl
    classoption: [onecolumn, oneside, a4paper]
    linkcolor: blue
    urlcolor: blue
    filecolor: magenta
    citecolor: magenta
    colorlinks: true
    margin-left: "1in"
    margin-right: "1in"
    margin-top: "1in"
    margin-bottom: "1in"
    papersize: a4
    fig-cap-location: top
    toc: true
    toc-depth: 5
    number-sections: true
    lof: false
    lot: false
link-citations: false
csl: "./literature/apa.csl"
bibliography: ["./literature/essential_readings.bib", "./literature/further_readings.bib"]
suppress-bibliography: true
execute:
  enabled: true
  echo: false
editor: 
  markdown: 
    wrap: 72
---

\clearpage

# Scope and Nature of Data Science

Let's start this course with some definitions and context.

**Definition of Data Science:**

> The ﬁeld of Data Science concerns techniques for extracting knowledge
> from diverse data, with a particular focus on ‘big’ data exhibiting
> ‘V’ attributes such as volume, velocity, variety, value and veracity.

@Maneth2016

**Definition of Data Analytics:**

Data analytics is the systematic process of examining data using
statistical, computational, and domain-specific methods to extract
insights, identify patterns, and support decision-making. It combines
competencies in data handling, analysis techniques, and domain knowledge
to generate actionable outcomes in organizational contexts
[@Cuadrado-Gallego2023].

**Definition of Business Analytics:**

> Business analytics is the science of posing and answering data
> questions related to business. Business analytics has rapidly expanded
> in the last few years to include tools drawn from statistics, data
> management, data visualization, and machine learn- ing. There is
> increasing emphasis on big data handling to assimilate the advances
> made in data sciences. As is often the case with applied
> methodologies, business analytics has to be soundly grounded in
> applications in various disciplines and business verticals to be
> valuable. The bridge between the tools and the applications are the
> modeling methods used by managers and researchers in disciplines such
> as finance, marketing, and operations.

@Pochiraju2019

There are many roles in the data science field, including (but not
limited to):

![**Source:**
*LinkedIn*](./topics/Scope%20and%20Nature%20of%20Data%20Science/IMG_0120.png)

For skills and competencies required for data science activities, see
[Skills
Landscape](./topics/Scope%20and%20Nature%20of%20Data%20Science/The%20data%20science%20skills%20landscape%20slides.pdf).

## Defining Data Science as an Academic Discipline

Data science emerges as an interdisciplinary field that synthesizes
methodologies and insights from multiple academic domains to extract
knowledge and actionable insights from data. As an academic discipline,
data science represents a convergence of computational, statistical, and
domain-specific expertise that addresses the growing need for
data-driven decision-making in various sectors.

Data science draws from and interacts with multiple foundational
disciplines:

-   **Informatics / Information Systems:**

    Informatics provides the foundational understanding of information
    processing, storage, and retrieval systems that underpin data
    science infrastructure. It encompasses database design, data
    modeling, information architecture, and system integration
    principles essential for managing large-scale data ecosystems.
    Information systems contribute knowledge about organizational data
    flows, enterprise architectures, and the sociotechnical aspects of
    data utilization in business contexts.

    See the [Technical Applications & Data Analytics
    coursebook](https://github.com/DrBenjamin/Data-Science-and-Data-Analytics/blob/66da3d5a65a1f57ab9ca5e35fd91df4077a7bad7/literature/Technical%20Applications%20%26%20Data%20Analytics.pdf/?raw=true)
    by @Gross2021 for further reading on foundations in informatics.

-   **Computer Science (algorithms, data structures, systems design):**

    Computer science provides the computational foundation for data
    science through algorithm design, complexity analysis, and efficient
    data structures. Core contributions include machine learning
    algorithms, distributed computing paradigms, database systems, and
    software engineering practices. System design principles enable
    scalable data processing architectures, while computational thinking
    frameworks guide algorithmic problem-solving approaches essential
    for data-driven solutions.

    See also: [Analytical Skills for Business - 1
    Introduction](https://drbenjamin.github.io/analytical-skills.html#sec-intro)
    and take a look at the AI Universe overview graphic:

    ![**Source:**
    *LinkedIn*](./topics/Scope%20and%20Nature%20of%20Data%20Science/IMG_0122.png)

    See the [Overview on no-code and low-code tools for data
    analytics](https://drbenjamin.github.io/analytical-skills.html#overview-on-no-code-and-low-code-tools-for-data-analytics)
    for an overview on no-code and low-code tools for data analytics and
    AI tooling.

-   **Mathematics (linear algebra, calculus, optimization):**

    Mathematics provides the theoretical backbone for data science
    through linear algebra (matrix operations, eigenvalues, vector
    spaces), calculus (derivatives, gradients, optimization), and
    discrete mathematics (graph theory, combinatorics). These
    mathematical foundations enable dimensionality reduction techniques,
    gradient-based optimization algorithms, statistical modeling, and
    the rigorous formulation of **machine learning problems** (see the
    figure below). Mathematical rigor ensures the validity and
    interpretability of analytical results.

    ![**Source:**
    *LinkedIn*](./topics/Scope%20and%20Nature%20of%20Data%20Science/AI_Models.png)

-   **Statistics & Econometrics (inference, modeling, causal
    analysis):**

    Statistics provides the methodological framework for data analysis
    through hypothesis testing, confidence intervals, regression
    analysis, and experimental design. Econometrics contributes advanced
    techniques for causal inference, time series analysis, and handling
    observational data challenges such as endogeneity and selection
    bias. These disciplines ensure rigorous uncertainty quantification,
    model validation, and the ability to draw reliable conclusions from
    data while understanding limitations and assumptions.

-   **Social Science & Behavioral Sciences (contextual interpretation,
    experimental design):**

    Social and behavioral sciences contribute essential understanding of
    human behavior, organizational dynamics, and contextual factors that
    influence data generation and interpretation. These disciplines
    provide expertise in experimental design, survey methodology,
    ethical considerations, and the social implications of data-driven
    decisions. They ensure that data science applications consider human
    factors, cultural context, and societal impact while maintaining
    ethical standards in data collection and analysis.

    ![**Source:**
    *LinkedIn*](topics/./Scope%20and%20Nature%20of%20Data%20Science/AI%20and%20culture.png)

The interdisciplinary nature of data science requires practitioners to
develop competencies across these domains while maintaining awareness of
how different methodological traditions complement and inform each
other. This multidisciplinary foundation enables data scientists to
approach complex problems with both technical rigor and contextual
understanding, ensuring that analytical solutions are both technically
sound and practically relevant.

For further reading on the academic foundations of data science, see the
comprehensive analysis in [Defining Data Science as an Academic
Discipline](./topics/Scope%20and%20Nature%20of%20Data%20Science/Defining%20Data%20Science%20as%20an%20Academic%20Discipline/482-1-945-1-10-20150421.pdf).

## Significance of Business Data Analysis for Decision-Making

Business data analysis has evolved from a supporting function to a
critical strategic capability that fundamentally transforms how
organizations make decisions, allocate resources, and compete in modern
markets. The systematic application of analytical methods to business
data enables evidence-based decision-making that reduces uncertainty,
improves operational efficiency, and creates sustainable competitive
advantages.

### Strategic Decision-Making Framework

Business data analysis provides a structured approach to strategic
decision-making through multiple analytical dimensions:

-   **Evidence-Based Strategic Planning**: Data analysis supports
    long-term strategic decisions by providing empirical evidence about
    market trends, competitive positioning, and organizational
    capabilities. Statistical analysis of historical performance data,
    market research, and competitive intelligence enables organizations
    to formulate strategies grounded in quantifiable evidence rather
    than intuition alone.

-   **Risk Assessment and Mitigation**: Advanced analytical techniques
    enable comprehensive risk evaluation across operational, financial,
    and strategic dimensions. Monte Carlo simulations, scenario
    analysis, and predictive modeling help organizations quantify
    potential risks and develop contingency plans based on probabilistic
    assessments of future outcomes.

-   **Resource Allocation Optimization**: Data-driven resource
    allocation models leverage optimization algorithms and statistical
    analysis to maximize return on investment across different business
    units, projects, and initiatives. Linear programming, integer
    optimization, and multi-criteria decision analysis provide
    frameworks for allocating limited resources to achieve optimal
    organizational outcomes.

### Operational Decision Support

At the operational level, business data analysis transforms day-to-day
decision-making through real-time insights and systematic performance
measurement:

-   **Performance Measurement and Continuous Improvement**: Key
    Performance Indicators (KPIs) and statistical process control
    methods enable organizations to monitor operational efficiency,
    quality metrics, and customer satisfaction in real-time. Time series
    analysis, control charts, and regression analysis identify trends,
    anomalies, and improvement opportunities that drive continuous
    operational enhancement.

-   **Forecasting and Demand Planning**: Statistical forecasting models
    using techniques such as ARIMA, exponential smoothing, and machine
    learning algorithms enable accurate demand prediction for inventory
    management, capacity planning, and supply chain optimization. These
    analytical approaches reduce uncertainty in operational planning
    while minimizing costs associated with overstock or stockouts.

-   **Customer Analytics and Personalization**: Advanced customer
    analytics leverage segmentation analysis, predictive modeling, and
    behavioral analytics to understand customer preferences, predict
    churn, and optimize retention strategies. Clustering algorithms,
    logistic regression, and recommendation systems enable personalized
    customer experiences that increase satisfaction and loyalty.

### Tactical Decision Integration

Business data analysis bridges strategic planning and operational
execution through tactical decision support:

-   **Pricing Strategy Optimization**: Price elasticity analysis,
    competitive pricing models, and revenue optimization techniques
    enable dynamic pricing strategies that maximize profitability while
    maintaining market competitiveness. Regression analysis, A/B
    testing, and econometric modeling provide empirical foundations for
    pricing decisions.

-   **Market Intelligence and Competitive Analysis**: Data analysis
    transforms market research and competitive intelligence into
    actionable insights through statistical analysis of market trends,
    customer behavior, and competitive positioning. Multivariate
    analysis, factor analysis, and time series forecasting identify
    market opportunities and competitive threats.

-   **Financial Performance Analysis**: Financial analytics encompassing
    ratio analysis, variance analysis, and predictive financial modeling
    enable organizations to assess financial health, identify cost
    reduction opportunities, and optimize capital structure decisions.
    Statistical analysis of financial data supports both internal
    performance evaluation and external stakeholder communication.

### Contemporary Analytical Capabilities

Modern business data analysis capabilities extend traditional analytical
methods through integration of advanced technologies and methodologies:

-   **Real-Time Analytics and Decision Support**: Stream processing,
    event-driven analytics, and real-time dashboards enable immediate
    response to changing business conditions. Complex event processing
    and real-time statistical monitoring support dynamic decision-making
    in fast-paced business environments.

-   **Predictive and Prescriptive Analytics**: Machine learning
    algorithms, neural networks, and optimization models enable
    organizations to not only predict future outcomes but also recommend
    optimal actions. These advanced analytical capabilities support
    automated decision-making and strategic scenario planning.

-   **Data-Driven Innovation**: Analytics-driven innovation leverages
    data science techniques to identify new business opportunities,
    develop innovative products and services, and create novel revenue
    streams. Advanced analytics enable organizations to discover hidden
    patterns, correlations, and insights that drive innovation and
    competitive differentiation.

The significance of business data analysis for decision-making extends
beyond technical capabilities to encompass organizational
transformation, cultural change, and strategic competitive positioning.
Organizations that successfully integrate analytical capabilities into
their decision-making processes achieve superior performance outcomes,
enhanced agility, and sustainable competitive advantages in increasingly
data-driven markets.

For comprehensive coverage of business data analysis methodologies and
applications, see [Advanced Business
Analytics](./topics/Scope%20and%20Nature%20of%20Data%20Science/Business%20Data%20Analysis/978-3-031-78070-7.pdf)
and the analytical foundations outlined in @Evans2020. 

For open access resources, visit [Kaggle](https://www.kaggle.com/),
a platform for data science competitions and datasets.

## Emerging Trends

Key technological and methodological developments shaping the data
landscape:

-   Evolution of computing and data processing architectures.

-   Digitalization of processes and platforms.

-   Artificial Intelligence (AI), Machine Learning (ML), and Deep
    Learning (DL).

-   Big Data ecosystems (volume, velocity, variety, veracity, value).

    ![**Source:**
    *LinkedIn*](./topics/Scope%20and%20Nature%20of%20Data%20Science/Emerging%20Trends/Big_Data_5Vs.jpg)

-   Internet of Things (IoT) and sensor-driven data generation.

    ![**Source:**
    *https://businesstech.bus.umich.edu/uncategorized/tech-101-internet-of-things/*](./topics/Scope%20and%20Nature%20of%20Data%20Science/Emerging%20Trends/iot2.png)

-   Cloud computing and elastic infrastructure.

-   Blockchain for distributed trust and data integrity.

-   Industry 4.0: cyber-physical systems and automation.

-   Remote and hybrid working environments: collaboration, distributed
    analytics, governance.

## Types of Analytics

-   Descriptive Analytics: What happened?
-   Predictive Analytics: What is likely to happen?
-   Prescriptive Analytics: What should we do?

![**Source:**
*https://datamites.com/blog/descriptive-vs-predictive-vs-prescriptive-analytics/*](./topics/Scope%20and%20Nature%20of%20Data%20Science/Emerging%20Trends/analytics.jpg)

\newpage

# Data Analytic Competencies

Data analytic competencies encompass the ability to apply machine
learning, data mining, statistical methods, and algorithmic approaches
to extract meaningful patterns, insights, and predictions from complex
datasets. They include proficiency in exploratory data analysis, feature
engineering, model selection, evaluation, and validation. These skills
ensure rigorous interpretation of data, support evidence-based
decision-making, and enable the development of robust analytical
solutions adaptable to diverse health, social, and technological
contexts.

## Types of Data

The structure and temporal dimension of data fundamentally influence
analytical approaches and statistical methods. Understanding data types
enables researchers to select appropriate modeling techniques and
interpret results within proper contextual boundaries.

-   **Cross-sectional data** captures observations of multiple entities
    (individuals, firms, countries) at a single point in time. This
    structure facilitates comparative analysis across units but does not
    track changes over time. Cross-sectional studies are particularly
    valuable for examining relationships between variables at a specific
    moment and testing hypotheses about population characteristics.

-   **Time-series data** records observations of a single entity across
    multiple time points, enabling the analysis of temporal patterns,
    trends, seasonality, and cyclical behaviors. Time-series methods
    account for autocorrelation and temporal dependencies, supporting
    forecasting and dynamic modeling. This data structure is essential
    for economic indicators, financial markets, and environmental
    monitoring.

-   **Panel (longitudinal) data** combines both dimensions, tracking
    multiple entities over time. This structure offers substantial
    analytical advantages by controlling for unobserved heterogeneity
    across entities and modeling both within-entity and between-entity
    variation. Panel data methods support causal inference through
    fixed-effects and random-effects models, difference-in-differences
    estimation, and dynamic panel specifications.

![**Source:**
*https://static.vecteezy.com*](./topics/Data%20Analytic%20Competencies/Types%20of%20Data/types%20of%20data.jpg)

**Additional data structures:**

-   **Geo-referenced / spatial data** is data associated with specific
    geographic locations, enabling spatial analysis and visualization.
    Techniques such as Geographic Information Systems (GIS), spatial
    autocorrelation, and spatial regression models are employed to
    analyze patterns and relationships in spatially distributed data.

![**Source:**
*https://www.slingshotsimulations.com/*](./topics/Data%20Analytic%20Competencies/Types%20of%20Data/GIS-Data.png)

-   **Streaming / real-time data** is continuously generated data that
    is processed and analyzed in real-time. This data structure is
    crucial for applications requiring immediate insights, such as fraud
    detection, network monitoring, and real-time recommendation systems.

## Types of Variables

-   **Continuous (interval/ratio)** data is measured on a scale with
    meaningful intervals and a true zero point (ratio) or arbitrary zero
    point (interval). Examples include height, weight, temperature, and
    income. Continuous variables support a wide range of statistical
    analyses, including regression and correlation.

```{r r_continuous, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading the dplyr library for data manipulation
library(dplyr)

# Assigning continuous data to a data frame and displaying it as a table
continuous_data <- data.frame(
  Height_cm = c(170, 165, 180, 175, 160, 185, 172, 168, 178, 182),
  Weight_kg = c(70, 60, 80, 75, 55, 90, 68, 62, 78, 85)
)

# Displaying the original table
print(continuous_data)

# Ordering the data frame by Height_cm
ordered_data <- continuous_data %>% 
  arrange(Height_cm)

# Displaying the ordered data frame
print(ordered_data)

# Practicing:
# 1. Assign continuous data to a data frame and displaying it as a table
# 2. Order the data frame by a specific column
```

-   **Count** data represents the number of occurrences of an event or
    the frequency of a particular characteristic. Count variables are
    typically non-negative integers and can be analyzed using Poisson
    regression or negative binomial regression.

```{r r_count, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Assigning count data to a data frame and displaying it as a table
count_data <- data.frame(
  Height = c(170, 165, 182, 175, 165, 175, 175, 168, 175, 182),
  Weight = c(70, 60, 80, 75, 55, 90, 68, 62, 78, 85)
)

# Displaying the original data as a table
print(count_data)

# Ordering the data frame by Height_cm in ascending order
ordered_count_data <- count_data %>%
  arrange(desc(Height), Weight) %>%
  count(Height)

# Displaying the ordered count data
print(ordered_count_data)

# Practicing:
# 1. Assign count data to a data frame and displaying it as a table
```

-   **Ordinal** data represents categories with a meaningful order or
    ranking but no consistent interval between categories. Examples
    include survey responses (e.g., Likert scales) and socioeconomic
    status (e.g., low, medium, high). Ordinal variables can be analyzed
    using non-parametric tests or ordinal regression.

```{r r_ordinal, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading the dplyr library for data manipulation
library(dplyr)

# Assigning ordinal data to a data frame and displaying it as a table
ordinal_data <- data.frame(
  Response = c("Strongly Disagree", "Disagree", "Neutral", "Agree", "Strongly Agree"),
  Value = c(5, 4, 3, 2, 1)
)

# Displaying the original table
print(ordinal_data)

# Ordering the data frame by Value
ordinal_data <- ordinal_data %>% 
  arrange(Value)

# Displaying the ordered data frame
print(ordinal_data)

# Practicing:
# 1. Assign ordinal data to a data frame and displaying it as a table
# 2. Order the data frame by the ordinal value
```

-   **Categorical (nominal / binary)** data represents distinct
    categories without any inherent order. Nominal variables have two or
    more categories (e.g., gender, race, or marital status), while
    binary variables have only two categories (e.g., yes/no,
    success/failure). Categorical variables can be analyzed using
    chi-square tests or logistic regression.

```{r r_categorical, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading the dplyr library for data manipulation
library(dplyr)

# Assigning categorical data to a data frame and displaying it as a table
categorical_data <- data.frame(
  Event = c("A", "C", "B", "D", "E"),
  Category = c("Type1", "Type2", "Type1", "Type3", "Type2")
)

# Displaying the original table
print(categorical_data)

# Ordering the data frame by Event
categorical_data <- categorical_data %>% 
  arrange(Event)

# Displaying the table
print(categorical_data)

# Practicing:
# 1. Assign categorical data to a data frame and displaying it as a table
# 2. Order the data frame by a specific column
```

-   **Compositional or hierarchical structures** represent data with a
    part-to-whole relationship or nested categories. Examples include
    demographic data (e.g., age groups within gender) and geographical
    data (e.g., countries within continents). Compositional data can be
    analyzed using techniques such as hierarchical clustering or
    multilevel modeling.

```{r r_compositional, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading the dplyr library for data manipulation
library(dplyr)

# Assigning hierarchical data to a data frame and displaying it as a table
hierarchical_data <- data.frame(
  Country = c("USA", "Canada" , "USA", "Canada", "Mexico"),
  State_Province = c("California", "Ontario", "Texas", "Quebec", "Jalisco"),
  Population_Millions = c(39.5, 14.5, 29.0, 8.5, 8.3)
)

# Displaying the table
print(hierarchical_data)

# Ordering the data frame by Country and then State_Province
hierarchical_data <- hierarchical_data %>% 
  arrange(Country, State_Province)

# Displaying the ordered data frame
print(hierarchical_data)

# Grouping the data by Country and summarizing total population
hierarchical_data_grouped <- hierarchical_data %>% 
  group_by(Country) %>%
  summarise(Total_Population = sum(Population_Millions))

# Displaying the grouped data frame
print(hierarchical_data_grouped)

# Practicing:
# 1. Assign hierarchical data to a data frame and displaying it as a table
# 2. Order the data frame by multiple columns
```

![**Source:**
*https://www.collegedisha.com/*](./topics/Data%20Analytic%20Competencies/Types%20of%20Data/Types_of_Data.png)

**Some small datasets to start with**

A custom R package `ourdata` has been created to provide some small datasets (and also some helper R functions) for practice. You can install it from GitHub using the following commands:

```{r r_package_inst, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Installing Github R packages
devtools::install_github("DrBenjamin/ourdata")
```

Using the package and exploring its documentation:

```{r r_packages_use, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading package
library(ourdata)

# Opening help of package
??ourdata

# Showing welcome message
ourdata()

# Printing some datasets
print(koelsch)
print(kirche)

# Using help function from the package for a specific function from the `ourdata` R package
help(combine)

# Using the `combine` function from the `ourdata` R package to combine two vectors into a data frame
ourdata::combine(kirche$Jahr, koelsch$Jahr, kirche$Austritte, koelsch$Koelsch)
```

** A messy dataset example**

We collect [data](https://data-explorer.oecd.org/vis?lc=en&df[ds]=dsDisseminateFinalDMZ&df[id]=DSD_HEALTH_STAT%40DF_AM&df[ag]=OECD.ELS.HD&dq=AUS%2BAUT%2BBEL%2BCAN%2BCHL%2BCOL%2BCRI%2BCZE%2BDNK%2BEST%2BFIN%2BFRA%2BDEU%2BGRC%2BHUN%2BISL%2BIRL%2BISR%2BITA%2BJPN%2BKOR%2BLVA%2BLTU%2BLUX%2BMEX%2BNLD%2BNZL%2BNOR%2BPOL%2BPRT%2BSVK%2BSVN%2BESP%2BSWE%2BCHE%2BTUR%2BGBR%2BUSA%2BARG%2BBRA%2BBGR%2BHRV%2BPER%2BROU%2BZAF%2BTHA.A.PREVM%2BTRTM.DT_10P5HB.._T...STANDARD....&pd=2010%2C2023&to[TIME_PERIOD]=false) from OECD (Organization for Economic Co-operation and Development), an international organization that works to build better policies for better lives. The dataset shows many columns (variables) with non-informative data and needs to be cleaned (wrangled) before analysis. First load the data into R from CSV file:

```{r oecd_data_1, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, eval=FALSE}
# Reading the dataset from a CSV file
preventable_deaths <- read.csv(
  "./topics/data/OECD_Preventable_Deaths.csv",
  stringsAsFactors = FALSE
)
```

or use the dataset directly from the `ourdata` R package:

```{r oecd_data_2, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Reading the dataset from `ourdata` R package
library(ourdata)
preventable_deaths <- oecd_preventable
```

First we explore the data:

```{r oecd_data_exp, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE, eval=FALSE}
# Viewing structure of the dataset
str(preventable_deaths)

# Viewing first few rows
head(preventable_deaths)

# Checking dimensions
dim(preventable_deaths)

# Viewing column names
colnames(preventable_deaths)

# Summary statistics
summary(preventable_deaths)

# Checking for missing values
colSums(is.na(preventable_deaths))
```

Now we can start cleaning the data by removing non-informative columns and rows with missing values:

```{r oecd_data_clean, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary library
library(dplyr)

# Selecting relevant columns for analysis
df_clean <- preventable_deaths %>%
  select(
    REF_AREA,
    Reference.area,
    TIME_PERIOD,
    OBS_VALUE
  ) %>%
  rename(
    country_code = REF_AREA,
    country = Reference.area,
    year = TIME_PERIOD,
    death_rate = OBS_VALUE
  )

# Converting year to numeric
df_clean$year <- as.numeric(df_clean$year)

# Converting death_rate to numeric (handling empty strings)
df_clean$death_rate <- as.numeric(df_clean$death_rate)

# Removing rows with missing death rates
df_clean <- df_clean %>%
  filter(!is.na(death_rate))

# Viewing cleaned data structure
str(df_clean)

# Summary of cleaned data
summary(df_clean)
```

Showing some basic statistics of the cleaned data:

```{r oecd_data_stats, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary library
library(dplyr)

# Printing overall statistics
cat("Mean death rate:", mean(df_clean$death_rate, na.rm = TRUE), "\n")
cat("Median death rate:", median(df_clean$death_rate, na.rm = TRUE), "\n")
cat("Standard deviation:", sd(df_clean$death_rate, na.rm = TRUE), "\n")
cat("Min death rate:", min(df_clean$death_rate, na.rm = TRUE), "\n")
cat("Max death rate:", max(df_clean$death_rate, na.rm = TRUE), "\n")

# Printing statistics by country
country_stats <- df_clean %>%
  group_by(country) %>%
  summarise(
    mean_rate = mean(death_rate, na.rm = TRUE),
    median_rate = median(death_rate, na.rm = TRUE),
    sd_rate = sd(death_rate, na.rm = TRUE),
    min_rate = min(death_rate, na.rm = TRUE),
    max_rate = max(death_rate, na.rm = TRUE),
    n_observations = n()
  ) %>%
  arrange(desc(mean_rate))
print(country_stats)

# Printing statistics by year
year_stats <- df_clean %>%
  group_by(year) %>%
  summarise(
    mean_rate = mean(death_rate, na.rm = TRUE),
    median_rate = median(death_rate, na.rm = TRUE),
    n_countries = n_distinct(country)
  ) %>%
  arrange(year)
print(year_stats)
```

To visualize the cleaned data, we can create a distribution plot of preventable death rates:

```{r oecd_data_viz, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary library
library(ggplot2)

# Plotting distribution of death rates
ggplot(df_clean, aes(x = death_rate)) +
  geom_histogram(bins = 30, fill = "#9B59B6", color = "white", alpha = 0.8) +
  labs(
    title = "Distribution of Preventable Death Rates",
    x = "Death Rate per 100,000",
    y = "Frequency",
    caption = "Source: OECD Health Statistics"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14)
  )
```

## Conceptual Framework: Knowledge & Understanding of Data

-   **Clarify analytical purpose and domain context** to guide data
    selection and interpretation.
-   **Define entities, observational units, and identifiers** to ensure
    accurate data representation.
-   **Align business concepts with data structures** for meaningful
    analysis.

## Data Collection

Data collection forms the foundational stage of any data science
project, requiring systematic approaches to gather information that
aligns with research objectives and analytical requirements. As outlined
in modern statistical frameworks, effective data collection strategies
must balance methodological rigor with practical constraints
[@etinkaya-Rundel2021].

![Methods of Data
Collection](./topics/Data%20Analytic%20Competencies/Data%20Collection/Method_of_data_collection.jpg)

### Core Data Collection Competencies

The competencies required for effective data collection encompass both
technical proficiency and methodological understanding (see [Data
Collection
Competencies.pdf](./topics/Data%20Analytic%20Competencies/Data%20Collection/Data%20Collection%20Competencies.pdf)):

-   **Source Identification and Assessment**: Systematically identify
    internal and external data sources, evaluating their relevance,
    quality, and accessibility for the analytical objectives.

-   **Data Acquisition Methods**: Implement appropriate collection
    techniques including APIs (for instance see [Spotify API tutorial](https://developer.spotify.com/documentation/web-api/howtos/web-app-profile) and [Postman Spotify tutorial](https://www.postman.com/gkoroscapidesign/spotify-api/collection/f03m16h/spotify-api)), database queries, survey instruments,
    sensor networks, web scraping, and third-party vendor partnerships,
    ensuring methodological alignment with research design.

    For the two projects below, here are some data sources recommendations (automatically created by
    Perplexity AI Deep Research):

    -   [Project car emission/sales](./topics/projects/car_emission_sales_data_scources.pdf)
    -   [Project Spotify](topics/projects/success_spotifiy_data_sources.pdf)

-   **Quality and Governance Framework**: Establish protocols for
    assessing data provenance, licensing agreements, ethical compliance,
    and regulatory requirements (GDPR, industry-specific standards).

-   **Methodological Considerations**: Apply principles from research
    methodology to ensure data collection approaches support valid
    statistical inference and minimize bias introduction during the
    acquisition process.

### Contemporary Data Collection Landscape

Modern data collection operates within an increasingly complex ecosystem
characterized by diverse data types, real-time requirements, and
distributed sources. The integration of traditional survey methods with
emerging IoT sensors, social media APIs, and automated data pipelines
requires comprehensive competency frameworks that address both technical
implementation and methodological validity.

*For comprehensive coverage of data collection methodologies and best
practices, refer to: [Research Methodology - Data
Collection](https://researchmethodology.org/data-collection/)*

## Data Management

-   Organize: schema design, naming conventions.
-   Clean: resolve duplicates, inconsistencies, missingness.
-   Convert: type casting, normalization, encoding.
-   Curate: maintain lineage, documentation, metadata.
-   Preserve: backups, versioning (see also [Analytical Skills for
    Business - Implementing version control
    systems](https://drbenjamin.github.io/analytical-skills.html#implementing-version-control-systems)),
    retention policies.

Data Management in data science curricula requires a coherent,
multi‑facet framework that spans data quality, FAIR stewardship, master
data governance, privacy/compliance, and modern architectures. Data
quality assessment and governance define objective metrics
(completeness, accuracy, consistency, plausibility, conformance) and
governance processes that balance automated checks with human oversight.
FAIR data principles provide a practical blueprint for metadata-rich
stewardship to support findability, accessibility, interoperability, and
reuse through machine-actionable metadata and persistent identifiers.
Master Data Management ensures clean, trusted core entities across
systems via governance and harmonization. Data privacy, security, and
regulatory compliance embed responsible data handling and risk
management, guided by purpose limitation, data minimization, accuracy,
storage limitations, integrity/confidentiality, and accountability.
Emerging trends in cloud-native data platforms, ETL/ELT, data
lakes/lakehouses, and broader metadata automation shape scalable
storage/compute and governance, enabling reproducible analytics and ML
workflows. Together, these strands underpin trustworthy, discoverable,
and compliant data inputs for research and coursework (Weiskopf & Weng,
2016; Wilkinson et al., 2016; GO FAIR Foundation, n.d.; Semarchy, n.d.;
IBM, n.d.).

## Data Evaluation

-   Define data quality dimensions and assessment frameworks.
-   Distinguish validation from verification in data pipelines.
-   Apply statistical methods and data profiling for evaluation.
-   Balance automated and manual (human-in-the-loop) evaluation
    approaches.
-   Implement tools, workflows, and governance for data evaluation.

Data evaluation ensures datasets are fit-for-purpose, reliable, and trustworthy throughout the analytical lifecycle. It encompasses systematic assessment of data quality dimensions (completeness, accuracy, consistency, validity, timeliness, uniqueness), validation and verification processes, and strategic application of statistical methods and data profiling. Quality dimensions provide measurable criteria for determining whether data meets analytical requirements, while assessment frameworks translate these into actionable metrics enabling objective measurement and contextual interpretation.

Validation and verification are complementary processes essential for data integrity. Validation checks occur at entry points, preventing bad data through constraint checks, format validation, and business rule enforcement. Verification involves post-collection checks ensuring data remains accurate and consistent over time and across system boundaries, supporting reproducibility and traceability. Together, validation acts as a gatekeeper while verification provides ongoing quality assurance.

Statistical methods form the technical foundation for evaluation. Outlier detection techniques (z-score, IQR, DBSCAN) identify anomalous observations requiring investigation. Distribution checks assess whether data conforms to expected patterns, while profiling describes dataset structure, missingness patterns, and statistical properties. Regression analysis and hypothesis testing diagnose quality issues and quantify relationships between quality metrics and analytical outcomes.

Modern data evaluation balances automated and manual approaches. Automated evaluation offers speed, scalability, and consistency for large datasets through rule-based validation, statistical profiling, and machine learning-based anomaly detection. Manual evaluation contributes domain expertise, contextual understanding, and interpretative judgment that automated systems cannot replicate. Human-in-the-loop approaches combine automation's efficiency with human interpretability, optimizing both throughput and quality.

Tools, workflows, and governance frameworks provide infrastructure for systematic evaluation across the data lifecycle. Data profiling tools (e.g., Pandas Profiling, Great Expectations, Deequ) automate quality assessment. Validation frameworks embed checks into ETL/ELT pipelines. Data lineage tracking and metadata management support traceability and impact analysis. Governance frameworks establish roles, responsibilities, and processes aligning evaluation practices with regulatory requirements and reproducibility needs.

\newpage

# Applications in the Programming Language R

Please read the [How to Use R for Data
Science](https://hubchev.github.io/ds/) by Prof. Dr. Huber for any basic
questions regarding R programming.

## Core tidyverse Tooling

Fundamental packages:

-   `dplyr` for data manipulation (filter, mutate, summarize, joins).

```{r dplyr_package, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary library
library(dplyr)

# Using dplyr to pipe
df_summary <- df_clean %>%
  group_by(country) %>%
  summarize(
    mean_death_rate = mean(death_rate, na.rm = TRUE),
    max_death_rate = max(death_rate, na.rm = TRUE),
    min_death_rate = min(death_rate, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_death_rate))

# Displaying the summary statistics
print(df_summary)
```

-   `tidyr` for data reshaping (pivoting, nesting, separating,
    unnesting).

```{r tidyr_package, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary library
library(tidyr)

# Using tidyr to pivot data
df_wide <- df_clean %>%
  pivot_wider(
    names_from = year,
    values_from = death_rate
  )

# Displaying the wide format data
print(df_wide)
```

-   `ggplot2` for layered grammar-based visualization. See the example above.

-   some more R libraries:

    -   `readr` for ingestion of CSV and other flat files.
    -   `readxl` for ingestion of Excel files.
    -   `lubridate` for date-time handling.
    -   `purrr` for functional iteration.
    -   `stringr` for text handling.
    -   `forcats` for factor handling.

    for ingestion, functional iteration, text, and factor handling.

## Data Visualization Principles

**Choose encodings appropriate to variable types:**

-   Continuous (Quantitative) variables → Position, Length
    Examples: x/y coordinates in scatter plots, bar heights, line positions
-   Categorical (Nominal) variables → Color hue, Shape, Facets
    Examples: different colors for groups, point shapes, separate panels
-   Ordinal variables → Ordered position, Color saturation
    Examples: ordered categories on axis, gradient colors from light to dark
-   Temporal variables → Position along x-axis, Line connections
    Examples: time on horizontal axis, connected points showing progression
-   Compositional (Part-to-whole) → Stacked position, Area
    Examples: stacked bars, proportional areas


**Emphasize clarity: reduce chart junk; apply perceptual best practices:**

Clarity in data visualization requires removing unnecessary elements that distract from the data while applying principles of human perception to enhance understanding.

-   Remove decorative elements (3D effects, shadows, gradients)
-   Eliminate redundant labels and gridlines
-   Minimize non-data ink (borders, background colors)
-   Avoid unnecessary legends when direct labeling is possible
-   Use position over angle for quantitative comparisons (bar charts > pie charts)
-   Maintain consistent scales across comparable charts
-   Respect aspect ratios that emphasize meaningful patterns
-   Choose colorblind-friendly palettes
-   Ensure sufficient contrast between data and background

```{r clarity_example, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary libraries
library(ggplot2)
library(dplyr)

# Creating data for demonstration
country_subset <- df_clean %>%
  filter(country %in% c("Germany", "France", "United Kingdom")) %>%
  filter(year >= 2015)

# Example: Clean, minimal visualization
ggplot(country_subset, aes(x = year, y = death_rate, color = country)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title = "Preventable Death Rates (2015-2023)",
    x = "Year",
    y = "Deaths per 100,000",
    color = "Country"
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),  # Removing unnecessary gridlines
    legend.position = "bottom",
    plot.title = element_text(face = "bold")
  )
```

**Support comparison, trend detection, and anomaly spotting:**

Effective visualizations should facilitate three key analytical tasks: comparing values across groups, identifying trends over time, and detecting unusual patterns.

**Support comparison:**

-   Align items on common scales for direct comparison
-   Use small multiples (facets) for comparing across categories
-   Order categorical variables meaningfully (by value, alphabetically, or logically)
-   Keep consistent ordering across related charts

**Enable trend detection:**

-   Use connected lines for temporal data to show continuity
-   Add trend lines (linear, loess) to highlight overall patterns
-   Display sufficient time periods to establish meaningful trends
-   Avoid over-smoothing that hides important variations

**Facilitate anomaly spotting:**

-   Use reference lines or bands for expected ranges
-   Highlight outliers through color or annotation
-   Include context (confidence intervals, historical ranges)
-   Maintain consistent scales to make deviations visible

```{r comparison_trend_example, echo=TRUE, message=FALSE, warning=FALSE, error=FALSE}
# Loading necessary libraries
library(ggplot2)
library(dplyr)

# Calculating statistics for anomaly detection
country_stats <- df_clean %>%
  group_by(country) %>%
  summarise(
    mean_rate = mean(death_rate, na.rm = TRUE),
    sd_rate = sd(death_rate, na.rm = TRUE)
  )

# Joining back to identify anomalies
df_annotated <- df_clean %>%
  left_join(country_stats, by = "country") %>%
  mutate(
    z_score = (death_rate - mean_rate) / sd_rate,
    is_anomaly = abs(z_score) > 2  # Flagging values > 2 standard deviations
  )

# Example: Visualization supporting comparison, trends, and anomaly detection
selected_countries <- c("Germany", "France", "United Kingdom", "Italy", "Spain")
df_subset <- df_annotated %>% filter(country %in% selected_countries)

ggplot(df_subset, aes(x = year, y = death_rate, color = country)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = is_anomaly, alpha = is_anomaly)) +
  scale_size_manual(values = c(1.5, 3), guide = "none") +
  scale_alpha_manual(values = c(0.6, 1), guide = "none") +
  facet_wrap(~ country, ncol = 2) +  # Small multiples for comparison
  labs(
    title = "Preventable Death Rates: Trends and Anomalies",
    subtitle = "Larger points indicate statistical anomalies (>2 SD from country mean)",
    x = "Year",
    y = "Deaths per 100,000",
    color = "Country"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## Detecting Outliers and Anomalies

-   Rule-based methods (IQR, z-scores).
-   Robust statistics (median, MAD).
-   Model-based or multivariate detection (e.g., Mahalanobis distance,
    clustering residuals).
-   Distinguish errors vs. novel but valid observations.

## Dimensionality Reduction

-   Motivation: mitigate multicollinearity, noise, and curse of
    dimensionality.
-   Techniques: Principal Component Analysis (PCA), Factor Analysis,
    (optionally) t-SNE / UMAP (for exploration).
-   Interpretability vs. compression trade-offs.

## Data Exploration and Mining

-   Structured EDA workflow: question → visualize → quantify → refine.
-   PCA for variance structure.
-   Factor Analysis for latent constructs.
-   Regression Analysis for relationships and predictive structure.
-   Clustering (k-means, hierarchical) for pattern discovery (if
    included).

## Causal Inference with Regression Analysis

-   Distinguish association vs. causation.
-   Model specification and confounding control.
-   Assumptions: linearity, independence, homoskedasticity, exogeneity.
-   Interpretation of coefficients and marginal effects.
-   Sensitivity and robustness checks.

\newpage

# Literature {#sec-literature}

All references for this course.

## Essential Readings

```{r essential_bib, results='asis', echo=FALSE, warning=FALSE}
source("literature/refmanager.r")
render_bib("literature/essential_readings.bib")
```

## Further Readings

```{r further_bib, results='asis', echo=FALSE, warning=FALSE}
source("literature/refmanager.r")
render_bib("literature/further_readings.bib")
```
